---
title: "Practical Machine Learning Course Project"
author: "Sanchit Sharma"
date: "13 December 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The goal of this project is to predict the manner in which they did the exercise.

## Loading required packages

``` {r pack}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```


## Load training and testing data

```{r load}
#loading the training data
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")

#loading the testing data
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")

#Read the training data and replace empty values by NA
trainingData <- read.csv("pml-training.csv", sep=",", header=TRUE, na.strings = c("NA","",'#DIV/0!'))
testingData <- read.csv("pml-testing.csv", sep=",", header=TRUE, na.strings = c("NA","",'#DIV/0!'))
```

## Looking at the data

```{r look}
#structure of training data
str(trainingData)

#structure of testing data
str(testingData)
```

Our data consists of 19622 values of 160 variables.



## Cleaning the data

We partition our training data into two

```{r part}
partTrain <- createDataPartition(trainingData$classe, p=0.6, list=FALSE)
myTrainData <- trainingData[partTrain, ]
myTestData <- trainingData[-partTrain, ]
dim(myTrainData); dim(myTestData)
```

To clean the data, we 
1.)Remove variables having zero variance.  2.)Remove the first column of myTrainData and myTestData data sets.  3.)Remove variables with more than 60% missing values.

```{r clean}
#Remove zero variance variables
novar <- nearZeroVar(myTrainData, saveMetrics=TRUE)
myTrainData <- myTrainData[,novar$nzv==FALSE]

novar <- nearZeroVar(myTestData, saveMetrics=TRUE)
myTestData <- myTestData[,novar$nzv==FALSE]

#Remove the first column of the myTrainData data set
myTrainData <- myTrainData[c(-1)]

#Clean variables with more than 60% missing values
trainData <- myTrainData
for(i in 1:length(myTrainData)) {
    if( sum( is.na( myTrainData[, i] ) ) /nrow(myTrainData) >= .7) {
        for(j in 1:length(trainData)) {
            if( length( grep(names(myTrainData[i]), names(trainData)[j]) ) == 1)  {
                trainData <- trainData[ , -j]
            }   
        } 
    }
}

# Set back to the original variable name
myTrainData <- trainData
rm(trainData)

#Transform the myTestData and testingData data sets
cleanData1 <- colnames(myTrainData)
cleanData2 <- colnames(myTrainData[, -58])  # remove the classe column
myTestData <- myTestData[cleanData1]         # allow only variables in myTestData that are also in myTrainData
testingData <- testingData[cleanData2]             # allow only variables in testingData that are also in myTrainData

dim(myTestData)

#Coerce the data into the same type
for (i in 1:length(testingData) ) {
  for(j in 1:length(myTrainData)) {
    if( length( grep(names(myTrainData[i]), names(testingData)[j]) ) == 1)  {
      class(testingData[j]) <- class(myTrainData[i])
    }      
  }      
}

# To get the same class between testing and myTraining
testingData <- rbind(myTrainData[2, -58] , testingData)
testingData <- testingData[-1,]



```

## Prediction with decision trees

```{r dt}
set.seed(3)
dtmodel <- rpart(classe ~ ., data=myTrainData, method="class")
fancyRpartPlot(dtmodel)

```

```{r sm}
predictionsdt <- predict(dtmodel, myTestData, type = "class")
cmtree <- confusionMatrix(predictionsdt, myTestData$classe)
cmtree
```


```{r p3}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))

```


## Prediction using Random Forest

```{r rf}

set.seed(1)
rfmodel <- randomForest(classe ~ ., data=myTrainData)
prediction1 <- predict(rfmodel, myTestData, type = "class")
cm <- confusionMatrix(prediction1, myTestData$classe)
cm

```

```{r graph}
plot(rfmodel)
```

```{r agraph}
plot(cm$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cm$overall['Accuracy'], 4)))
```


## Prediction with generalization booster

```{r gbm}
set.seed(2)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmmodel <- train(classe ~ ., data=myTrainData, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)


gbmFinModel <- gbmmodel$finalModel

gbmPrediction <- predict(gbmmodel, newdata=myTestData)
gbmAccuracy <- confusionMatrix(gbmPrediction, myTestData$classe)
gbmAccuracy

```

```{r plot}
plot(gbmmodel, ylim=c(0.9, 1))
```

## Predicting results on test data

Random Forests gave an Accuracy in the myTesting dataset of 99.82%, which was more accurate that what I got from the Decision Trees or GBM. The expected out-of-sample error is 100-99.82 = 0.18%.

```{r test}
prediction2 <- predict(rfmodel, testingData, type = "class")
prediction2

```
